This paper explored the potential of NVSHMEM for enhancing the performance of communication-intensive workloads such as shuffling operations in distributed database systems.
Using NVSHMEM, we were able to leverage GPU-initated communication - as opposed to the traditional CPU-initiated approach - for optimized data transfers.

We made several key findings and contributions in this paper.
Using NVSHMEM, we built a GPU-initiated shuffle algorithm and benchmarked it, evaluating multiple algorithmic options.

Our microbenchmarks demonstrate that while NVSHMEM can achieve reasonable throughput, it also exhibits performance characteristics that are non-trivial to understand and do not align with initial expectations.

Additionally, we found out that while the use of NVSHMEM can reduce the kernel launch overhead, it comes at the cost of a more complex programming model, requiring careful design for communication between the host and device.

While NVSHMEM shows clear potential for performance improvements, it is not clear whether this approach is appropriate for general database applications.
In future work, a deeper dive into the intricacies of NVSHMEM's internal workings is warranted to explore its usefulness in a distributed database context.


% Points to mentions: investigated performance aspects of nvshmem by means of microbenchmarks. Implemented a shuffling algorithm with different approaches for sending and synchronization. Nvshmem was not designed for use in a database environment which makes adaption complicated.